---
title: "Electric Power Consumption & CO₂ Emissions Forecasting"
subtitle: "Time Series Analysis | Europe & Central Asia | 1981–2023"
author: "Tracy Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

---

# Load Libraries

```{r libraries}
library(bookdown)
library(tidyverse)
library(ggplot2)
library(forecast)
library(TTR)
library(dplyr)
library(tidyr)
library(tseries)
```

---

# Load Data

```{r load-data}
setwd("/Users/thuthao/Documents/STUDY/SUSS/ANL557/ECA")
df <- read.csv("Word Bank Data_All.csv")
```

---

# Reshape Dataset

> **Goal:** Transform the raw wide-format World Bank data into two clean univariate
> time series — one for Electric Power Consumption, one for CO₂ Emissions.

## Separate Each Series

Filter each series from the combined dataset into its own data frame.

```{r separate-series}
electric_data2 <- df %>%
  filter(Series.Name == "Electric power consumption (kWh per capita)") %>%
  select(Country.Name, starts_with("X"))

co2_data2 <- df %>%
  filter(Series.Name == "Carbon dioxide (CO2) emissions from Waste (Mt CO2e)") %>%
  select(Country.Name, starts_with("X"))
```

## Convert Columns to Character Before Pivoting

Year columns must be character type to avoid type conflicts during `pivot_longer`.

```{r convert-character}
electric_data2 <- electric_data2 %>%
  mutate(across(starts_with("X"), as.character))

co2_data2 <- co2_data2 %>%
  mutate(across(starts_with("X"), as.character))
```

## Pivot to Long Format

Reshape from wide (one column per year) to long (one row per year),
then clean the Year column and convert values to numeric.

```{r pivot-long}
electric_data_long2 <- electric_data2 %>%
  pivot_longer(cols = starts_with("X"),
               names_to  = "Year",
               values_to = "Electric_Power_Consumption") %>%
  mutate(Year = gsub("^X|\\.\\..YR", "", Year)) %>%  # Remove "X" prefix and "..YR" suffix
  mutate(Year = substr(Year, 1, 4)) %>%              # Keep only the 4-digit year
  mutate(Year = as.integer(Year)) %>%                # Convert to integer
  mutate(Electric_Power_Consumption = na_if(Electric_Power_Consumption, "..")) %>%  # Replace ".." with NA
  mutate(Electric_Power_Consumption = as.numeric(Electric_Power_Consumption))

co2_data_long2 <- co2_data2 %>%
  pivot_longer(cols = starts_with("X"),
               names_to  = "Year",
               values_to = "CO2_Emissions") %>%
  mutate(Year = gsub("^X|\\.\\..YR", "", Year)) %>%
  mutate(Year = substr(Year, 1, 4)) %>%
  mutate(Year = as.integer(Year)) %>%
  mutate(CO2_Emissions = na_if(CO2_Emissions, "..")) %>%
  mutate(CO2_Emissions = as.numeric(CO2_Emissions))
```

## Combine Both Series

Join Electric Power and CO₂ into a single data frame by Country and Year.

```{r combine-data}
combined_data <- left_join(electric_data_long2, co2_data_long2,
                           by = c("Country.Name", "Year"))
str(combined_data)
```

---

# Data Cleaning

## Identify Missing Values

```{r missing-values}
arr.miss_electric <- is.na(combined_data$Electric_Power_Consumption)
arr.miss_co2      <- is.na(combined_data$CO2_Emissions)  # Expected: no missing values

cat("Missing — Electric Power Consumption:", sum(arr.miss_electric), "\n")
cat("Missing — CO2 Emissions             :", sum(arr.miss_co2), "\n")
```

## Detect Outliers Using the IQR Rule

Values outside `[Q1 − 1.5×IQR, Q3 + 1.5×IQR]` are flagged as outliers.

```{r outlier-detect}
# IQR bounds for Electric Power Consumption
iqr_bounds_electric <- quantile(combined_data$Electric_Power_Consumption,
                                probs = c(0.25, 0.75), na.rm = TRUE) +
  c(-1.5, 1.5) * IQR(combined_data$Electric_Power_Consumption, na.rm = TRUE)

# IQR bounds for CO2 Emissions
iqr_bounds_co2 <- quantile(combined_data$CO2_Emissions,
                           probs = c(0.25, 0.75), na.rm = TRUE) +
  c(-1.5, 1.5) * IQR(combined_data$CO2_Emissions, na.rm = TRUE)

# Flag outliers in both series
combined_data$outlier_electric <- with(combined_data,
  Electric_Power_Consumption < iqr_bounds_electric[1] |
  Electric_Power_Consumption > iqr_bounds_electric[2])

combined_data$outlier_co2 <- with(combined_data,
  CO2_Emissions < iqr_bounds_co2[1] |
  CO2_Emissions > iqr_bounds_co2[2])
```

## Impute Missing Values via Linear Regression

Since CO₂ Emissions has no missing values and is strongly correlated with
Electric Power Consumption, we use a linear regression model to predict
and fill in missing electricity values.

**Step 1 — Check correlation to justify this approach:**

```{r correlation-check}
correlation <- cor(combined_data$Electric_Power_Consumption,
                   combined_data$CO2_Emissions, use = "complete.obs")
cat("Correlation coefficient (pre-imputation):", round(correlation, 3), "\n")
```

**Step 2 — Visualise the relationship:**

```{r scatter-before-imputation}
ggplot(combined_data, aes(x = Electric_Power_Consumption, y = CO2_Emissions)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title    = "Electric Power Consumption vs CO₂ Emissions",
    subtitle = "Pre-imputation — confirming correlation before regression-based fill",
    x = "Electric Power Consumption (kWh per capita)",
    y = "CO₂ Emissions (Mt CO2e)"
  ) +
  theme_minimal()
```

**Step 3 — Fit the model and replace missing values:**

```{r imputation}
# Build regression model on complete observations only
lm_model <- lm(Electric_Power_Consumption ~ CO2_Emissions, data = combined_data)

# Identify rows where Electric Power is missing
missing_idx <- which(is.na(combined_data$Electric_Power_Consumption))

# Replace missing values with model predictions
combined_data$Electric_Power_Consumption[missing_idx] <-
  predict(lm_model, newdata = combined_data[missing_idx, ])
```

**Step 4 — Visualise the time series after imputation:**

```{r plot-after-imputation}
ggplot(combined_data, aes(x = Year, y = Electric_Power_Consumption)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "black") +
  labs(
    title    = "Electric Power Consumption (kWh per Capita)",
    subtitle = "Europe & Central Asia, 1981–2023 — after imputation",
    x = "Year",
    y = "kWh per Capita"
  ) +
  theme_minimal()
```

## Re-check Outliers After Imputation

Recalculate IQR bounds on the now-complete dataset and visualise using
a boxplot and histogram.

```{r outlier-after}
iqr_bounds_electric_after <- quantile(combined_data$Electric_Power_Consumption,
                                      probs = c(0.25, 0.75), na.rm = TRUE) +
  c(-1.5, 1.5) * IQR(combined_data$Electric_Power_Consumption, na.rm = TRUE)

combined_data$outlier_electric_after <- with(combined_data,
  Electric_Power_Consumption < iqr_bounds_electric_after[1] |
  Electric_Power_Consumption > iqr_bounds_electric_after[2])
```

```{r outlier-boxplot}
ggplot(combined_data, aes(x = "", y = Electric_Power_Consumption)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.shape = NA) +
  geom_jitter(aes(color = outlier_electric_after), width = 0.2, size = 2) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  labs(
    title    = "Boxplot of Electric Power Consumption After Imputation",
    subtitle = "Outliers highlighted in red",
    x = NULL,
    y = "kWh per Capita"
  ) +
  theme_minimal()
```

```{r outlier-histogram}
ggplot(combined_data, aes(x = Electric_Power_Consumption)) +
  geom_histogram(binwidth = 200, fill = "lightblue", color = "black") +
  labs(
    title    = "Histogram of Electric Power Consumption After Imputation",
    subtitle = "Distribution from 1981 to 2023",
    x = "kWh per Capita",
    y = "Frequency"
  ) +
  theme_minimal()
```

---

# Time Series Analysis

## Visualise the Full Time Series

```{r ts-visualise}
ggplot(combined_data, aes(x = Year, y = Electric_Power_Consumption)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "black") +
  labs(
    title    = "Electric Power Consumption (kWh per Capita)",
    subtitle = "Europe & Central Asia, 1981–2023",
    x = "Year",
    y = "kWh per Capita"
  ) +
  theme_minimal()
```

## Convert to Time Series Object

```{r ts-object}
combined_data <- combined_data[order(combined_data$Year), ]
ts_electric <- ts(combined_data$Electric_Power_Consumption,
                  start = 1981, end = 2023, frequency = 1)
```

## Check for Seasonality (ACF)

```{r acf-seasonality}
acf(combined_data$Electric_Power_Consumption, lag.max = 24, na.action = na.pass)
```

> **Finding:** ACF shows a slow, gradual decay without repeating spikes — indicating
> a **trend** component but **no clear seasonality**. Analysis will focus on trend
> and irregular components only.

## Trend Analysis — Linear Regression

```{r trend-model}
combined_data$Index <- 1:nrow(combined_data)

lm_trend <- lm(Electric_Power_Consumption ~ Index, data = combined_data)
summary(lm_trend)
```

> **Finding:** The model confirms a statistically significant upward trend of
> approximately **21 kWh per capita per year** (R² = 0.685, p < 0.001).

## Residuals (Irregular Component)

Extract and visualise the residuals to understand the irregular component
after removing the linear trend.

```{r residuals-extract}
combined_data$Residuals <- lm_trend$residuals
```

```{r residuals-plot}
ggplot(combined_data, aes(x = Year, y = Residuals)) +
  geom_line(color = "darkred") +
  labs(
    title = "Residuals from Linear Trend (Irregular Component)",
    x = "Year",
    y = "Residuals"
  ) +
  theme_minimal()
```

```{r residuals-histogram}
ggplot(combined_data, aes(x = Residuals)) +
  geom_histogram(fill = "steelblue", bins = 10) +
  labs(
    title = "Histogram of Residuals (Irregularities)",
    x = "Residual",
    y = "Frequency"
  ) +
  theme_minimal()
```

---

# Forecasting Models

> **Setup:** Models are trained on **1981–2022** data. The actual 2023 value
> (5,536 kWh) is reserved as a hold-out to evaluate forecast accuracy.

## Prepare Training Data

```{r training-data}
subset_data <- subset(combined_data, Year >= 1981 & Year <= 2022)
ts_data <- ts(subset_data$Electric_Power_Consumption,
              start = 1981, end = 2022, frequency = 1)

fc_step          <- 2  # Steps for accuracy evaluation
forecast_horizon <- 2  # Forecast 2023 and 2024
```

## Exponential Smoothing Models

```{r exp-smoothing}
# Simple Exponential Smoothing — good for series with no trend or seasonality
ses_model <- ses(ts_data, h = fc_step)
ses_fc    <- ses_model

# Holt's Linear Trend (Double Exponential Smoothing) — captures trend
holt_model <- holt(ts_data, h = fc_step)
holt_fc    <- holt_model
```

## Stationarity Check (Required Before MA / ARIMA)

MA and ARIMA models require a stationary series. We use the ADF test
and ACF/PACF plots to verify.

```{r stationarity}
adf.test(combined_data$Electric_Power_Consumption)
```

```{r acf-pacf-original}
acf(ts_data,  main = "ACF — Original Series")
pacf(ts_data, main = "PACF — Original Series")
```

> **Result:** ADF p-value > 0.05 → series is **non-stationary**.
> First differencing is required before fitting MA/ARIMA models.

## First Differencing

```{r differencing}
ts_diff <- diff(ts_data)

acf(ts_diff,  main = "ACF — Differenced Series")
pacf(ts_diff, main = "PACF — Differenced Series")
```

```{r ar1-check}
# Fit AR(1) to differenced series to inspect the autoregressive coefficient
ar1_model <- Arima(ts_diff, order = c(1, 0, 0))
summary(ar1_model)
```

## Moving Average Models

```{r moving-average}
# MA(1): ARIMA(0,1,1) — one lagged error term after differencing
ma1_model <- Arima(ts_data, order = c(0, 1, 1))
ma1_fc    <- forecast(ma1_model, h = fc_step, level = 95)

# MA(2): ARIMA(0,1,2) — two lagged error terms after differencing
ma2_model <- Arima(ts_data, order = c(0, 1, 2))
ma2_fc    <- forecast(ma2_model, h = fc_step, level = 95)
```

## ARIMA Model

```{r arima}
# auto.arima() automatically selects the best p, d, q parameters
arima_model <- auto.arima(ts_data)
arima_fc    <- forecast(arima_model, h = fc_step, level = 95)
```

## Accuracy Comparison

Compare all five models using RMSE, MAE, and MAPE on the training set.

```{r accuracy}
accuracy(ma1_fc)
accuracy(ma2_fc)
accuracy(arima_fc)
accuracy(ses_fc)
accuracy(holt_fc)
```

---

# Forecast for 2023 and 2024

## Check Actual 2023 Value

```{r actual-2023}
# Retrieve the actual 2023 observation from the original dataset
combined_data[combined_data$Year == 2023, ]
```

## Generate Forecasts from All Models

```{r generate-forecasts}
# MA(1)
ma1_forecast <- forecast(ma1_model, h = forecast_horizon)
ma1_fc_pred  <- ma1_forecast$mean
ma1_fc_cil   <- ma1_forecast$lower[, "95%"]
ma1_fc_ciu   <- ma1_forecast$upper[, "95%"]

# MA(2)
ma2_forecast <- forecast(ma2_model, h = forecast_horizon)
ma2_fc_pred  <- ma2_forecast$mean
ma2_fc_cil   <- ma2_forecast$lower[, "95%"]
ma2_fc_ciu   <- ma2_forecast$upper[, "95%"]

# SES
ses_forecast <- forecast(ses_model, h = forecast_horizon)
ses_fc_pred  <- ses_forecast$mean

# Holt's Linear Trend
holt_forecast <- forecast(holt_model, h = forecast_horizon)
holt_fc_pred  <- holt_forecast$mean

# ARIMA
arima_forecast <- forecast(arima_model, h = forecast_horizon)
arima_fc_pred  <- arima_forecast$mean
arima_fc_cil   <- arima_forecast$lower[, "95%"]
arima_fc_ciu   <- arima_forecast$upper[, "95%"]
```

## Visualise — Holt's Linear Trend Forecast

```{r plot-holt}
years_holt  <- 1981:2024
values_holt <- c(ts_data, as.numeric(holt_fc_pred))

holt_df <- data.frame(
  Year  = years_holt,
  Value = values_holt,
  Type  = c(rep("Actual", length(ts_data)), rep("Forecast", length(holt_fc_pred)))
)

ggplot(holt_df, aes(x = Year, y = Value, color = Type)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  labs(
    title    = "Electric Power Consumption Forecast — Holt's Linear Trend",
    subtitle = "Forecast for 2023–2024 using Holt's Exponential Smoothing",
    x = "Year",
    y = "kWh per capita"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "darkgreen")) +
  theme_minimal()
```

## Visualise — MA(2) Forecast

```{r plot-ma2}
years_ma2  <- 1981:2024
values_ma2 <- c(ts_data, as.numeric(ma2_fc_pred))

ma2_df <- data.frame(
  Year  = years_ma2,
  Value = values_ma2,
  Type  = c(rep("Actual", length(ts_data)), rep("Forecast", length(ma2_fc_pred)))
)

ggplot(ma2_df, aes(x = Year, y = Value, color = Type)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  labs(
    title    = "Electric Power Consumption Forecast — MA(2)",
    subtitle = "Forecast for 2023–2024 using Moving Average (order 2)",
    x = "Year",
    y = "kWh per capita"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "red")) +
  theme_minimal()
```

---

# Relationship: Electric Power vs CO₂ Emissions

## Correlation Coefficient

```{r correlation-final}
correlation <- cor(combined_data$Electric_Power_Consumption,
                   combined_data$CO2_Emissions, use = "complete.obs")
print(paste("Correlation coefficient:", round(correlation, 3)))
```

> A value of **r = 0.829** indicates a **strong positive relationship** —
> as electricity consumption increases, CO₂ emissions rise proportionally.
> This confirms electricity consumption as a reliable proxy for emissions tracking.

## Scatter Plot

```{r scatter-final}
ggplot(combined_data, aes(x = Electric_Power_Consumption, y = CO2_Emissions)) +
  geom_point() +
  labs(
    title = "Scatter Plot: Electric Power Consumption vs CO₂ Emissions",
    x     = "Electric Power Consumption (kWh per capita)",
    y     = "CO₂ Emissions (Mt CO2e)"
  ) +
  theme_minimal()
```

## Dual-Axis Time Series Plot

Since CO₂ values are much smaller in scale than electricity consumption,
a scaling factor is applied so both series can be plotted on the same chart
with a secondary axis for readability.

```{r ts-dual-axis}
scaling_factor <- 1000

ggplot(combined_data, aes(x = Year)) +
  geom_line(aes(y = Electric_Power_Consumption,
                color = "Electric Power Consumption")) +
  geom_line(aes(y = CO2_Emissions * scaling_factor,
                color = "CO2 Emissions")) +
  scale_y_continuous(
    name      = "Electric Power Consumption (kWh per capita)",
    sec.axis  = sec_axis(~ . / scaling_factor,
                         name = "CO2 Emissions (metric tons per capita)")
  ) +
  scale_color_manual(values = c(
    "Electric Power Consumption" = "blue",
    "CO2 Emissions"              = "red"
  )) +
  labs(
    title = "Electric Power Consumption vs CO₂ Emissions Over Time",
    x     = "Year"
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())
```

---

*Analysis by Tracy Nguyen | ANL557 Applied Forecasting | SUSS 2025*
